---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.retina = 3
)
```

# cbcTools

<!-- badges: start -->
[![CRAN
status](https://www.r-pkg.org/badges/version/cbcTools)](https://CRAN.R-project.org/package=cbcTools)
<!-- badges: end -->

This package contains tools for designing surveys and conducting power analyses for choice based conjoint survey experiments in R.

## Installation

The current version is not yet on CRAN, but you can install it from
Github using the {remotes} library:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("jhelvy/cbcTools")
```

Load the library with:
```{r}
library("cbcTools")
```

## Make experiment designs

The first step in designing an experiment is to define the attributes and levels for your experiment. Many of the functions in {cbcTools} are more convenient to use if you define these as a separate object. For example, let's say you're designing a conjoint experiment about apples. You might have the following attributes and levels:

```{r}
# Define the attributes and levels
levels <- list(
  price     = seq(1, 4, 0.5), # $ per pound
  type      = c('Fuji', 'Gala', 'Honeycrisp', 'Pink Lady', 'Red Delicious'),
  freshness = c('Excellent', 'Average', 'Poor')
)
```

With these levels defined, making the design of experiment is straightforward using the `makeDoe()` function:

```{r}
doe <- makeDoe(levels)
head(doe)
```

You can also make a fractional-factorial design based on different criteria. For example, to make a "D-optimal" design, add the `type = "D"` argument and specify the number of trials you want to use from the full factorial design (you can also use `"A"` and `"I"` to optimize for different efficiency measures):

```{r}
doe <- makeDoe(levels, type = "D", nTrials = 50)
```

If you're unsure of how many trials to use, you can set `search = TRUE` to search across all feasible designs up to `nTrials` (with an optional `minTrials` argument as a starting number of trials in the search):

```{r}
doe <- makeDoe(levels, type = "D", nTrials = 50, minTrials = 30, search = TRUE)
```

If using `search = TRUE`, a summary of all designs will be printed to the console, and the design with the highest D-efficiency is returned.

You can also check the D-efficiency of any design using the `evaluateDoe()` function:

```{r}
evaluateDoe(doe)
```

Once you've made your design, you can easily re-code it using the actual labels in your `levels` object using `recodeDoe()`:

```{r}
doe <- recodeDoe(doe, levels)
head(doe)
```

You should re-code your design prior to making a survey if you want to show the actual label values in the survey rather than just the attribute level numbers (e.g., 1, 2, 3, etc.).

## Make conjoint surveys

A survey can be generated by randomly sampling from the design of experiment. The `makeSurvey()` function also eliminates the possibility of duplicate alternatives appearing in the same choice question:

```{r}
survey <- makeSurvey(
    doe       = doe,  # Design of experiment
    nResp     = 2000, # Total number of respondents (upper bound)
    nAltsPerQ = 3,    # Number of alternatives per question
    nQPerResp = 6     # Number of questions per respondent
)

dim(survey)
head(survey)
```

The resulting data frame includes the following additional columns:

- `respID`: Identifies each survey respondent.
- `qID`: Identifies the choice question answered by the respondent.
- `altID`:Identifies the alternative in any one choice observation.
- `obsID`: Identifies each unique choice observation across all respondents.

You can also make a "labeled" survey where the levels of one of the attributes is used as a label by setting the `group` argument to that attribute. This by definition sets the number of alternatives in each question to the number of levels in the chosen attribute. Here is an example labeled survey using the "freshness" attribute as the label:

```{r}
survey_labeled <- makeSurvey(
    doe       = doe,
    nResp     = 2000,
    nAltsPerQ = 3,
    nQPerResp = 6,
    group     = "freshness"
)

dim(survey_labeled)
head(survey_labeled)
```

In the above example, you can see in the first six rows of the survey that the `freshness` attribute is always fixed to be the same order, ensuring that each level in the `freshness` attribute will always be shown in each choice question.

Finally, you can also include an "Outside Good" (e.g., a "none" option) in your survey by setting `outsideGood = TRUE`. If included, all categorical attributes are dummy-coded to appropriately dummy-code the outside good alternative.

```{r}
survey_og <- makeSurvey(
    doe       = doe,
    nResp     = 2000,
    nAltsPerQ = 3,
    nQPerResp = 6,
    outsideGood = TRUE
)

dim(survey_og)
head(survey_og)
```

## Simulate choices

You can simulate choices for a given `survey` using the `simulateChoices()` function. By default, random choices are simulated:

```{r}
data <- simulateChoices(
    survey = survey,
    obsID  = "obsID"
)
head(data)
```

You can also pass a list of parameters to define a utility model that will be used to simulate choices. In the example below, the choices are simulated using a utility model with the following parameters:

- 1 continuous `price` parameter
- 4 discrete parameters for `type`
- 2 discrete parameters for `freshness`

```{r, eval=FALSE}
data <- simulateChoices(
    survey = survey,
    obsID  = "obsID",
    pars = list(
        price     = 0.1,
        type      = c(0.1, 0.2, 0.3, 0.4),
        freshness = c(0.1, -0.1))
)
```

You can also simulate data with more complex models, such as mixed logit models where parameters follow a normal or log-normal distribution across the population, or interaction between parameters. In the example below, the choices are simulated using a utility model with the following parameters:

- 1 continuous "price" parameter
- 4 discrete parameters for "type"
- 2 random normal discrete parameters for "freshness"
- 2 interaction parameters between "price" and "freshness"

The `randN()` function is use to make the 2 `freshness` parameters follow a normal distribution with a specified mean (`mu`) and standard deviation (`sigma`).

```{r, eval=FALSE}
data <- simulateChoices(
    survey = survey,
    obsID  = "obsID",
    pars = list(
        price     = 0.1,
        type      = c(0.1, 0.2, 0.3, 0.4),
        freshness = randN(mu = c(0.1, -0.1), sigma = c(1, 2)),
        `price*freshness` = c(1, 2))
)
```

## Conduct a power analysis

The simulated choice data can be used to conduct a power analysis by estimating multiple models with different sample sizes. The `estimateModels()` function achieves this by partitioning the simulated choice data into multiple sizes (defined by the `nbreaks` argument) and then estimating a user-defined choice model on each data subset. In the example below, 10 different sample sizes are used to estimate 10 models.

```{r}
models <- estimateModels(
    nbreaks = 10,
    data    = data,
    pars    = c("price", "type", "freshness"),
    outcome = "choice",
    obsID   = "obsID"
)
```

The resulting `models` object is a list of estimated models, each estimated using the [{logitr}](https://jhelvy.github.io/logitr) package.

While the `models` object is a rather complex object in that it contains multiple models, helper functions can be used to extract information of interest. For example, the estimated coefficients and standard errors from each model can be extracted using the `getModelResults()` function:

```{r}
results <- getModelResults(models)

head(results)
```

Here is a summary of the standard errors for each sample size:

```{r}
library(ggplot2)

ggplot(results) +
  geom_hline(yintercept = 0.05, color = "red", linetype = 2) +
  geom_point(aes(x = sampleSize, y = se, color = coef)) +
  expand_limits(y = 0) +
  theme_bw()
```

## Author, Version, and License Information

- Author: *John Paul Helveston* https://www.jhelvy.com/
- Date First Written: *October 23, 2020*
- License: [MIT](https://github.com/jhelvy/cbcTools/blob/master/LICENSE.md)

## Citation Information

If you use this package for in a publication, I would greatly appreciate it if you cited it - you can get the citation by typing `citation("cbcTools")` into R:

```{r}
citation("cbcTools")
```
